---
title: "Class 12"
author: "Trinity Lee A16639698"
format: pdf
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo=TRUE)
```

#section 1. Proportions of G/G in a population 
downloaded a CSV file from Ensemble and read CSV
```{r}
mxl<-read.csv("373531-SampleGenotypes-Homo_sapiens_variation_Sample_rs8067378.csv")
head(mxl)

```

```{r}
table(mxl$Genotype..forward.strand.)
table(mxl$Genotype..forward.strand.)/nrow(mxl)*100
```

Now lets look at a different population. I picked GBR. 
```{r}
gbr<-read.csv("373522-SampleGenotypes-Homo_sapiens_variation_Sample_rs8067378.csv")
head(gbr)
```
Find proportion of G/G

```{r}
table(gbr$Genotype..forward.strand.)
round(table(gbr$Genotype..forward.strand.)/nrow(gbr)*100,2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MXL population 

Now let's dig into this further.

#4 population analysis
```{r}
expr<-read.table("sample geno.txt")
head(expr)
```

Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes
```{r}
nrow(expr)
table(expr$geno)

```


```{r}
library(ggplot2)

```

lets make a boxplot

```{r}
ggplot(expr)+aes(x=geno,y=exp,fill=geno)+geom_boxplot(notch=TRUE)
```
what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3? 
G/G does have a significant impact on the expression of the ORMDL3 being less compared to genotype A/A as seen by the difference in box plots. 
