---
title: "class 05 data visualization with ggplot2"
author: "Trinity Lee A16639698"
format: pdf
---

## Using GGPLOT

The ggplot2 package does not already come installed with R

We have to use `install.packages()` function to install ggplot2

```{r}
head(cars)
```

To use ggplot I need to load it up before I can call any of the functions in the package. I do this with the `library()` function. 

```{r}
library(ggplot2)
ggplot()
```

All ggplot figures have at least three things: 
- data (the stuff we want to plot)
- aesthetic mapping (aes vales)
- geoms

```{r}
ggplot(cars)+aes(x=speed,y=dist) + geom_point()
```

ggplot is not the only graphing system in R there are lots of others. THere is even "base R" graphics.

```{r}
plot(cars)
```

```{r}
ggplot(cars)+aes(x=speed,y=dist) + geom_point() + geom_smooth()
```

```{r}
ggplot(cars)+aes(x=speed,y=dist) + geom_point() + geom_smooth(method="lm",se=FALSE)
```

```{r}
ggplot(cars)+aes(x=speed,y=dist) + geom_point() + geom_smooth(method="lm",se=FALSE)+labs(title="speed vs stopping distance of cars", x="speed of cars (mph)",y="stopping distance (ft)")+theme_bw()
```

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```
```{r}
nrow(genes)
```
```{r}
colnames(genes)
ncol(genes)
```
```{r}
table(genes$State)
table(genes$State)/nrow(genes)*100
```
```{r}
p<-ggplot(genes)+aes(x=Condition1, y=Condition2, col=State)+geom_point()
p+scale_color_manual (values=c("red","orange","yellow"))+labs(title="Drug treatment influence on Gene Expression", x= "no drug used (control)",y="treatment with drug")
```

```{r}
# install.packages("dplyr")  ## un-comment to install if needed
library(dplyr)
library(gapminder)
gapminder_2007 <- gapminder %>% filter(year==2007)
ggplot(gapminder_2007)+aes(x=gdpPercap, y=lifeExp, size=pop)+geom_point(alpha=0.5)+scale_size_area(max_size=10)
```
```{r}
gapminder_1957<-gapminder %>% filter(year==1957|year==2007)
ggplot(gapminder_1957)+aes(x=gdpPercap, y=lifeExp, size=pop,color=continent)+geom_point(alpha=0.5)+scale_size_area(max_size=10)+facet_wrap(~year)
```

```{r}
gapminder_top5 <- gapminder %>% 
  filter(year==2007) %>% 
  arrange(desc(pop)) %>% 
  top_n(5, pop)

gapminder_top5
ggplot(gapminder_top5)+aes(x=reorder(country,-pop),y=pop,fill=country)+ geom_col(col="gray30")+guides(fill="none")
```

```{r}
USArrests$State <- rownames(USArrests)
ggplot(USArrests) +
  aes(x=reorder(State,Murder), y=Murder) +
  geom_point() +geom_segment(aes(x=State,xend=State,y=0,yend=Murder),color="red")+coord_flip()
```
```

